---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hermiteStats: Distribution-Robust Statistics via Hermite Polynomial Quantile Modeling

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/WLenhard/hermiteStats/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/WLenhard/hermiteStats/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**`hermiteStats`** provides distribution-robust, regularized estimators for distributional moments (mean, variance, skewness and kurtosis) and based on this, for two of the most fundamental statistics in quantitative research:

1. **The Hermite Correlation (r_Hermite):** a distribution-robust estimator of the raw-scale Pearson correlation coefficient.
2. **The Distribution-Free Effect Size (d_reg):** a robust standardized mean difference for independent and paired samples that retains the metric and interpretation of Cohen's *d* / Hedges' *g*.

Both methods address a long-standing dilemma in robust statistics: classical remedies for non-normal data — trimming, Winsorizing, or converting to ranks — typically restore stability only by silently changing the quantity being estimated, discarding genuine tail variance in the process. `hermiteStats` instead models the empirical quantile function via **regularized monotone polynomial smoothing** and derives exact distributional moments in closed algebraic form using **Probabilists' Hermite polynomials** and either without any prior assumptions, or - in case of the additional Gaussian copula assumption - based on **Mehler's (1866) identity**. It recovers the original, interpretable estimand while gaining the stability of a robust method.

---

## Table of Contents

- [Key Features](#key-features)
- [Mathematical Foundations](#mathematical-foundations)
  - [1. Rank-Based Inverse-Normal Quantile Modeling](#1-rank-based-inverse-normal-quantile-modeling)
  - [2. The Orthogonal Hermite Basis and Univariate Moments](#2-the-orthogonal-hermite-basis-and-univariate-moments)
  - [3. Mehler's Bilinear Expansion for Covariance](#3-mehlers-bilinear-expansion-for-covariance)
  - [4. Disentangling Construct Association from Scale Attenuation (A)](#4-disentangling-construct-association-from-scale-attenuation-a)
  - [5. Distribution-Free Effect Size (d_reg)](#5-distribution-free-effect-size-d_reg)
- [Installation](#installation)
- [Quick Start and Examples](#quick-start-and-examples)
  - [Estimating Distributional Moments](#estimating-distributional-moments)
  - [Hermite-Mehler Correlation (r_Hermite)](#hermite-mehler-correlation-r_Hermite)
  - [Correlation Matrices and Multivariate Analysis](#correlation-matrices-and-multivariate-analysis)
  - [Distribution-Free Effect Sizes (d_reg)](#distribution-free-effect-sizes-d_reg)
  - [Paired and Repeated-Measures Designs](#paired-and-repeated-measures-designs)
  - [Diagnostic Visualizations](#diagnostic-visualizations)
- [Citation](#citation)
- [References](#references)

---

## Key Features

* **No Estimand Shift:** Retains the raw-scale Pearson correlation and standardized mean difference as the target parameters — unlike rank correlations or Winsorized effect sizes, which silently estimate a different population quantity.
* **Variance Reduction in Small Samples (n < 50):** Regularized, low-degree polynomial quantile smoothing prevents extreme tail observations from destabilizing sample covariances and pooled standard deviations.
* **Negligible Efficiency Loss under Normality:** Closely tracks the classical Pearson's *r* / Cohen's *d* when parametric assumptions hold, so there is essentially nothing to lose by using the regularized estimator as a default choice.
* **Flexible Monotonicity Enforcement:** Supports both a relaxed, rank-concordance-based check and a strict, analytically verified monotonicity constraint (root-finding on the fitted derivative).
* **Fully Closed-Form Computation:** All moments and covariances are obtained algebraically via Hermite/Mehler identities — no numerical integration, simulation, or iterative optimization is required.

---

## Mathematical Foundations

### 1. Rank-Based Inverse-Normal Quantile Modeling

Let $X$ and $Y$ be continuous random variables with unknown cumulative distribution functions $F_X$ and $F_Y$. The observed data are mapped to latent standard normal scores $Z \sim \mathcal{N}(0, 1)$ using a rank-based inverse-normal transformation (Normal Quantile Transformation; e.g. Hazen, 1914):

$$Z_{x, i} = \Phi^{-1}\left(\frac{\text{rank}(X_i) - 0.5}{n}\right), \quad Z_{y, i} = \Phi^{-1}\left(\frac{\text{rank}(Y_i) - 0.5}{n}\right)$$

Unlike a plain rank transform, this step is only a means to an end: the inverse quantile functions $X = f(Z_x)$ and $Y = g(Z_y)$ are then modeled by regressing the raw values on powers of $Z$,

$$X = \sum_{j=0}^{k_x} \beta_{x, j} Z_x^j, \quad Y = \sum_{j=0}^{k_y} \beta_{y, j} Z_y^j,$$

so that the original metric and tail shape of the data are preserved rather than discarded. To guarantee a valid quantile map, monotonicity of $f$ (and $g$) is enforced over the observed range of $Z$, automatically stepping down the polynomial degree whenever a violation is detected.

### 2. The Orthogonal Hermite Basis and Univariate Moments

Standard powers $Z^j$ are not orthogonal under the Gaussian measure, which makes their coefficients awkward for moment extraction. Re-expressing the polynomial in the basis of **monic Probabilists' Hermite polynomials** $He_m(z)$ (Hermite, 1864),

$$He_0(z) = 1, \quad He_1(z) = z, \quad He_2(z) = z^2 - 1, \quad He_3(z) = z^3 - 3z, \quad \dots$$

solves this: because $He_m(z)$ is orthogonal with respect to the standard normal density, $\mathbb{E}[He_m(Z) He_n(Z)] = m! \, \delta_{mn}$, the monomial coefficients $\boldsymbol{\beta}$ map to Hermite coefficients $\mathbf{a}$ via a fixed linear operator, $\mathbf{a} = \mathbf{H}\boldsymbol{\beta}$, with

$$H_{m+1, n+1} = \binom{n}{2k} (2k - 1)!! \quad \text{for } 2k = n - m \ge 0 \text{ (and } 0 \text{ otherwise)}.$$

The mean and variance then follow immediately from the Hermite coefficients:

$$\mathbb{E}[X] = a_0, \qquad \text{Var}(X) = \sum_{m=1}^{k_x} a_m^2 \, m!$$

Skewness $\gamma_1$ and excess kurtosis $\gamma_2$ follow analogously from the third and fourth central moments of the fitted polynomial, evaluated using the raw Gaussian moments $\mathbb{E}[Z^j]$ — a direct consequence of Isserlis' (1918) theorem for products of jointly normal variables:

$$\gamma_1 = \frac{\mathbb{E}[(X-\mu)^3]}{\sigma^3}, \qquad \gamma_2 = \frac{\mathbb{E}[(X-\mu)^4]}{\sigma^4} - 3$$

All four moments are obtained this way in exact closed algebraic form, with no numerical integration at any step.

### 3. Covariance Estimation on the Basis of Probabilists' Hermite Polynomials

Two different variants for calculating covariance and thus the resulting $r_{\text{HM}}$ are available:

#### Variant A: Copula-Free Empirical Cross-Moments (`copula = "none"`, Default)

The first variant is based directly on the fitted raw scores ($\hat{x}_i, \hat{y}_i$) and the regularized Hermite population moments. It is completely free of prior assumptions regarding the joint dependence structure (copula) and marginal shape. It yields lower MSE and higher precision across a vast range of distributional scenarios (especially in sample sizes of $n < 100$):

$$\widehat{\text{Cov}}(X, Y) = \frac{\sum_{i=1}^n \left(\hat{x}_i - \hat{\mu}_x\right)\left(\hat{y}_i - \hat{\mu}_y\right)}{n}$$

where:
* $\hat{x}_i = f(Z_{x, i}) = \sum_{j=0}^{k_x} \beta_{x, j} Z_{x, i}^j$ is the fitted value for observation $i$ from the monotone quantile model.
* $\hat{\mu}_x = a_0$ is the regularized population mean derived from the Hermite polynomial representation.

Standardizing by the regularized marginal standard deviations ($\hat{\sigma}_x, \hat{\sigma}_y$) yields the **Copula-Free Hermite Correlation**:

$$r_{\text{Hermite}} = \frac{\widehat{\text{Cov}}(X, Y)}{\hat{\sigma}_x \, \hat{\sigma}_y} = \frac{\frac{1}{n} \sum_{i=1}^n \left(\hat{x}_i - \hat{\mu}_x\right)\left(\hat{y}_i - \hat{\mu}_y\right)}{\sqrt{\left(\sum_{m=1}^{k_x} a_m^2 \, m!\right)\left(\sum_{m=1}^{k_y} b_m^2 \, m!\right)}}$$

---

#### Variant B: Parametric Gaussian Copula via Mehler's Identity (`copula = "gaussian"`)

The second variant evaluates the cross-moment expectations analytically under the assumption that the latent normal scores $(Z_x, Z_y)$ follow a bivariate normal distribution with correlation $\rho_z = \text{cor}(Z_x, Z_y)$. By **Mehler's (1866) bilinear expansion**, all cross-orders of different degrees vanish ($\mathbb{E}[He_m(Z_x) He_n(Z_y)] = \delta_{mn} m! \rho_z^m$), yielding the closed-form covariance:

$$\widehat{\text{Cov}}_{\text{Gauss}}(X, Y) = \sum_{m=1}^{\min(k_x, k_y)} a_m b_m \, m! \, \rho_z^m$$

and the corresponding **Gaussian-Copula Hermite Correlation**:

$$r_{\text{Hermite, Gauss}} = \frac{\sum_{m=1}^{\min(k_x, k_y)} a_m b_m \, m! \, \rho_z^m}{\sqrt{\left(\sum_{m=1}^{k_x} a_m^2 \, m!\right)\left(\sum_{m=1}^{k_y} b_m^2 \, m!\right)}}$$

This variant also yields the **Shape Attenuation Factor** $A = r_{\text{Hermte}} / \rho_z$, which quantifies the mathematical ceiling imposed purely by marginal shape incompatibility (Hoeffding–Fréchet bounds).

### 4. Disentangling Construct Association from Scale Attenuation (A)

This framework cleanly separates two conceptually distinct quantities: the pure latent dependency $\rho_z$ (the Gaussian copula parameter, unaffected by either variable's shape), and the manifest linear correlation $r_{\text{Hermite}}$ (which *is* affected by shape, exactly as classical Pearson's $r$ would be). Their ratio defines the **Shape Attenuation Factor (A)**:

$$A = \frac{r_{\text{Hermite}}}{\rho_z}$$

* If $A \approx 1.0$, the marginal distributions do not meaningfully constrain the correlation.
* If $A \ll 1.0$, part of what looks like a "weak" Pearson correlation is in fact a mathematical ceiling imposed by mismatched marginal shapes (Hoeffding-Fréchet bounds; Carroll, 1961) rather than weak underlying dependence.

### 5. Distribution-Free Effect Size (d_reg)

For two independent groups, marginal quantile polynomials $X_1 = f_1(Z_1)$ and $X_2 = f_2(Z_2)$ are fitted separately to obtain regularized population moments $\hat{\mu}_1, \hat{\mu}_2, \hat{\sigma}_1^2, \hat{\sigma}_2^2$. The standardized mean difference is then

$$d_{\text{reg}} = \frac{\hat{\mu}_2 - \hat{\mu}_1}{\hat{\sigma}_{\text{avg}}}, \qquad \hat{\sigma}_{\text{avg}} = \sqrt{\frac{\hat{\sigma}_1^2 + \hat{\sigma}_2^2}{2}}$$

For **paired / repeated-measures designs** (`paired = TRUE`), the observed difference scores $D = X_2 - X_1$ are modeled directly with their own regularized quantile fit, giving the standardized mean change $d_z = \hat{\mu}_D / \hat{\sigma}_D$. This $\hat{\sigma}_D$ is theoretically linked to the marginal moments and the paired Hermite-Mehler correlation via the general variance-of-a-difference identity

$$\hat{\sigma}_D = \sqrt{\hat{\sigma}_1^2 + \hat{\sigma}_2^2 - 2\, r_{\text{Hermite}} \, \hat{\sigma}_1 \hat{\sigma}_2}$$

(no equal-variance assumption required), which makes explicit why a stronger paired correlation mechanically shrinks the variance of the change scores, and hence inflates $d_z$ relative to $d_{\text{reg}}$.

---

## Installation
`hermiteStats` is not yet on CRAN. Install the development version directly from GitHub:

```{r, eval = FALSE}
# install.packages(\"remotes\")
remotes::install_github(\"WLenhard/hermiteStats\")
```

The package depends only on base R and requires R >= 3.5.0 — there are no dependencies to resolve.


## Quick Start and Examples

```{r, eval = TRUE}
library(hermiteStats)
```

### Estimating Distributional Moments

At the core of the package is `hermite_fit()`, which maps a numeric vector onto standard normal scores, fits a regularized monotone polynomial quantile function, and extracts the population mean, variance, skewness, and excess kurtosis in exact closed algebraic form, without the instability of raw sample variance, skewness and kurtosis estimators in small or heavy-tailed samples.

```{r, eval = TRUE}
set.seed(42)
x <- rlnorm(40, meanlog = 3, sdlog = 0.5)  # right-skewed, heavy-tailed data

fit <- hermite_fit(x)
print(fit)
```


`hermite_moments()` returns the same regularized moments as a plain list, convenient for programmatic use:

```{r, eval = TRUE}
hermite_moments(fit)
```


`summary()` additionally shows the fitted monomial coefficients and the orthogonal Hermite weights the moments above were derived from:


```{r, eval = TRUE}
summary(fit)
```

For comparison, the raw sample mean and standard deviation on the same data:

```{r, eval = TRUE}
mean(x)
sd(x)
```

In a clean, moderately sized continuous sample like this one, the regularized and raw estimates are expected to agree closely — `hermite_fit()`'s advantage shows up primarily in small samples, tied or contaminated data, and whenever skewness/kurtosis are of direct interest, as the remaining examples below illustrate.



### Hermite Correlation (r_Hermite)

`cor_hermite()` estimates the raw-scale Pearson correlation between two variables that may have very different marginal shapes — here, a roughly normal construct and a skewed, lognormal latency, both driven by a shared latent factor:

```{r, eval = TRUE}
set.seed(42)
n <- 60
z_shared <- rnorm(n)

x <- 100 + 15 * z_shared + rnorm(n, sd = 5)      # approximately normal
y <- exp(0.4 * z_shared + rnorm(n, sd = 0.5))    # right-skewed (lognormal)

fit_cor <- cor_hermite(x, y, conf_level = 0.95, ci_method = "fisher")
print(fit_cor)
```

The Hermite correlation makes no assumption on the distributional shape and the copula (the type of bivariate dependence structure). If the copula is Gaussian - a decision to be made on the basis of theoretical considerations - this can be specified via `copula = "gauss"`, leading to a further decrease of MSE. In that case, further parameters are available: `rho_z` is the latent Gaussian copula correlation; `r_Hermite` is the corresponding raw-scale Pearson correlation implied by the fitted marginal shapes; and `A = r_Hermite / rho_z` quantifies how much of that latent association survives on the manifest scale (see [Mathematical Foundations §4](#4-disentangling-construct-association-from-scale-attenuation-a)). `summary(fit_cor)` additionally reports the implied means, standard deviations, and covariance on the raw scale.


### Correlation Matrices and Multivariate Analysis

For more than two variables, `cor_hermite()` dispatches to a matrix/data-frame method, returning all three quantities as full pairwise matrices — useful as a drop-in, shape-robust alternative to `cor()` ahead of factor analysis, SEM, or meta-analytic synthesis:

```{r, eval = TRUE}
R_hermite <- cor_hermite(iris[, 1:4])
print(R_hermite)
```

### Distribution-Free Effect Sizes (d_reg)

`d_reg()` estimates a standardized mean difference that keeps Cohen's *d* metric and interpretation, but is far less sensitive to skewness and small-sample instability. It accepts either two vectors or, as here, a formula interface:

```{r, eval = TRUE}
set.seed(123)
df_study <- data.frame(
  score = c(rlnorm(25, meanlog = 2.0, sdlog = 0.5),   # Control (skewed)
            rlnorm(25, meanlog = 2.4, sdlog = 0.5)),  # Treatment
  group = factor(rep(c("Control", "Treatment"), each = 25))
)

fit_d <- d_reg(score ~ group, data = df_study, conf_level = 0.95, ci_method = "bootstrap")
print(fit_d)
```

You can as well directly compare two variables with `d_reg(x, y)`. Alongside `d_reg`, the object also carries the classical Hedges' *g* benchmark and, on request, a percentile bootstrap confidence interval; `summary(fit_d)` reports the regularized moments underlying both groups.

### Paired and Repeated-Measures Designs

For within-subject designs, `paired = TRUE` fits the marginal distributions *and* the observed difference-score distribution, reporting both the raw-scale effect size and the standardized mean change:

```{r, eval = TRUE}
set.seed(7)
pre  <- rlnorm(30, meanlog = 3.0, sdlog = 0.4)
post <- pre + rnorm(30, mean = 5.0, sd = 2.0)

fit_paired <- d_reg(pre, post, paired = TRUE, conf_level = 0.95)
print(fit_paired)
```

`d_reg` (raw scale) is directly comparable to independent-groups effect sizes, e.g. in a meta-analysis mixing designs; `d_z` is standardized by the variability of the change scores themselves and will be larger whenever the paired Hermite correlation is strong (see [Mathematical Foundations §5](#5-distribution-free-effect-size-d_reg)).

### Diagnostic Visualizations

Every model object has a `plot()` method, so the regularization step is never a black box.

A single fitted quantile model shows the empirical normal scores against the raw data, overlaid with the fitted polynomial and its regularized moments:

```{r, eval = TRUE}
plot(fit)
```

A `cor_hermite` fit shows the latent copula on the left, and the manifest association on the right — with both the model-implied conditional mean curve and a purely empirical lowess smooth, so that curvature attributable to marginal shape can be visually distinguished from a genuine departure from the constant-correlation copula assumption:

```{r, eval = TRUE}
plot(fit_cor)
```

A `d_reg` fit shows both groups' empirical densities (with regularized means and the effect size annotated) alongside their regularized quantile maps overlaid on a single panel:

```{r, eval = TRUE}
plot(fit_d)
```


## Citation
If you use `hermiteStats` in published research, please cite both the software and the manuscripts describing the two estimators. Both manuscripts are currently under peer review; please check the package repository for updated publication details (journal, volume, DOI) once available.

```bibtex
@Manual{lenhard2026hermiteStatsPkg,
  title  = {{hermiteStats}: Distribution-Robust Statistics via Hermite Polynomial Quantile Modeling},
  author = {Wolfgang Lenhard and Alexandra Lenhard},
  year   = {2026},
  note   = {R package version 0.1.0},
  url    = {https://github.com/WLenhard/hermiteStats}
}
@Unpublished{lenhard2026hermite,
  title  = {The {Hermite} Correlation: A Distribution-Robust Estimator of the {Pearson} Correlation Coefficient},
  author = {Wolfgang Lenhard and Alexandra Lenhard},
  note   = {Manuscript submitted for publication}
}

@Unpublished{lenhard2026dreg,
  title  = {Distribution-Free Effect Size Estimation: A Robust Alternative to {Cohen's d} and Other Effect Size Estimators},
  author = {Wolfgang Lenhard and Alexandra Lenhard},
  note   = {Manuscript submitted for publication}
}
```

## References

* Carroll, J. B. (1961). The nature of the data, or how to choose a correlation coefficient. *Psychometrika*, 26(4), 347–372. https://doi.org/10.1007/BF02289768
* Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.). Lawrence Erlbaum Associates.
* Hazen, A. (1914). Storage to be provided in impounding municipal water supply. *Transactions of the American Society of Civil Engineers*, 77(1), 1539–1640.
* Hedges, L. V. (1981). Distribution theory for Glass's estimator of effect size and related estimators. *Journal of Educational Statistics*, 6(2), 107–128. https://doi.org/10.3102/10769986006002107
* Hermite, C. (1864). Sur un nouveau développement en série des fonctions. *Comptes Rendus de l'Académie des Sciences, Paris*, 58, 93–100.
* Isserlis, L. (1918). On a formula for the product-moment coefficient of any order of a normal frequency distribution. *Biometrika*, 12(1/2), 134–139. https://doi.org/10.1093/biomet/12.1-2.134
* Mehler, F. G. (1866). Ueber die Entwicklung einer Function von beliebig vielen Variabeln nach Laplaceschen Functionen höherer Ordnung. *Journal für die reine und angewandte Mathematik*, 66, 161–176. https://doi.org/10.1515/crll.1866.66.161
* Sklar, M. (1959). Fonctions de répartition à n dimensions et leurs marges. *Publications de l'Institut de Statistique de l'Université de Paris*, 8, 229–231.

---
